---
title: "Strawberry_EDA"
author: "Wenshuo Cai"
date: "2023-10-24"
output:
  pdf_document: default
  html_document: default
---
## Load the data

```{r setup, include=FALSE }
#| label: Load libraries
#| warning: false
#| message: false
#| echo: false

library(knitr)  
library(tidyverse)
library(stringr)
library(reshape2)
library(ggplot2)
library(dplyr)
library(tidyr)

```



```{r warning=FALSE, message=FALSE}
#| label: read data - glimpse 
#| warning: false
#| message: false
#| echo: true

strawberry <- read_csv("strawberry.csv", col_names = TRUE)

glimpse(strawberry)

```

## Data Cleaning 

```{r}
#| label: drop one-item columns
#| echo: false
# Convert the 'Value' column to numeric, handling non-numeric entries gracefully
strawberry$Value <- suppressWarnings(as.numeric(as.character(strawberry$Value)))

# Define a function to remove columns with a single unique value
remove_single_value_columns <- function(data_frame) {
  single_value_columns <- c()
  for (col in names(data_frame)) {
    if (length(unique(data_frame[[col]])) == 1) {
      single_value_columns <- c(single_value_columns, col)
    }
  }
  
  if (length(single_value_columns) == 0) {
    return("No single-value columns found")
  } else {
    removed_info <- data.frame(Column = single_value_columns, Value = sapply(single_value_columns, function(col) data_frame[1, col, drop = TRUE]))
    return(removed_info)
  }
}

# Apply the function to the dataframe
removed_columns_info <- remove_single_value_columns(strawberry)

# Display the names and values of the removed columns
print(removed_columns_info)

# If any single-value columns were found, remove them from the dataframe
if (is.data.frame(removed_columns_info)) {
  strawberry <- strawberry[, !(names(strawberry) %in% removed_columns_info$Column)]
}


```
```{r}
#| label: glimpse of strawberry data
#| echo: false

glimpse(strawberry)
```

###Examine the data.
```{r}
#| label: examine rows
#| echo: true
## is every line associated with a state?

## state_all1 contains the number of rows containing data 
## for each of the 47 strawberry-growing states.
state_all1 <- strawberry |> group_by(State) |> count()

## test if every row is associated with a state by summing the 
## counts and testing for equality with the total rows in the 
## data frame

if(sum(state_all1$n) == dim(strawberry)[1]){print("Every row has value in the State column.")}

```

```{r}
#| label: which state has the most rows
#| echo: true

state_max <- state_all1$State[which(state_all1$n ==  max(state_all1$n)  )]
print(state_max)
```


```{r}
#| label: examine California data
#| echo: true

## filter rows of California data from the CENSUS data
calif_census <- strawberry |> filter((State=="CALIFORNIA") & (Program=="CENSUS"))


## ## filter rows of California data from the SURVEY data
calif_survey <- strawberry |> filter((State=="CALIFORNIA") & (Program=="SURVEY"))

 
census_col <- colnames(calif_census)

survey_col <- colnames(calif_survey)


```


```{r}
#|label: split srawberry into census and survey pieces
#|echo:false

strwb_census <- strawberry |> filter(Program == "CENSUS")

strwb_survey <- strawberry |> filter(Program == "SURVEY")

## check that all of the rows are accounted for

nrow(strawberry) == (nrow(strwb_census) + nrow(strwb_survey))

## Move marketing-related rows in strw_b_chem 
## to strw_b_sales
strwb_census <- strwb_census |>
  separate_wider_delim(  cols = `Data Item`,
                         delim = ",",
                         names = c("Fruit",
                                 "temp1",
                                 "temp2",
                                 "temp3"),
                         too_many = "error",
                         too_few = "align_start"
                       )



## split temp1 into crop_type, Prop_acct

strwb_census <- strwb_census |>
  separate_wider_delim(  cols = temp1,
                         delim = " - ",
                         names = c("crop_type",
                                 "Prop_acct"),
                         too_many = "error",
                         too_few = "align_start"
                       )


# Define a function to clean and trim a column
clean_and_trim <- function(data_frame, column) {
  data_frame %>%
    mutate(!!column := str_trim(!!sym(column), side = "both")) %>%
    mutate(!!column := na_if(!!sym(column), "")) # Convert NA to empty string
}

# Define a function to remove specific patterns from a column
remove_patterns <- function(data_frame, column, patterns) {
  for (pattern in patterns) {
    data_frame <- data_frame %>%
      mutate(!!column := str_replace_all(!!sym(column), pattern, ""))
  }
  data_frame
}

# Clean and trim columns
strwb_census <- clean_and_trim(strwb_census, "crop_type")
strwb_census <- clean_and_trim(strwb_census, "temp2")
strwb_census <- clean_and_trim(strwb_census, "temp3")

# Handle `Fresh Market` column
strwb_census <- strwb_census %>%
  mutate(`Fresh Market` = temp2) %>%
  remove_patterns("Fresh Market", c("^MEA.*", "^P.*", "^FRESH MARKET - "))

# Handle `Process Market` column
strwb_census <- strwb_census %>%
  mutate(`Process Market` = temp2) %>%
  remove_patterns("Process Market", c("^MEA.*", "PROCESSING - "))

# Handle `temp2` column
strwb_census <- remove_patterns(strwb_census, "temp2", c("^F.*", "^P.*"))

# Combine `temp2` and `temp3`, clean `Metric` column, and relocate columns
strwb_census <- strwb_census %>%
  unite("Metric", c("temp2", "temp3"), sep = "") %>%
  mutate(Metric = str_replace(Metric, "MEASURED IN ", "")) %>%
  relocate(Metric, .before = Domain) %>%
  rename(Totals = Prop_acct)



```

```{r}
#| label: fix the leading space
#| echo: false

# Apply the trimming operation to the existing columns
strawberry$`Data Item` <- str_trim(strawberry$`Data Item`, side = "both")
strawberry$Domain <- str_trim(strawberry$Domain, side = "both")
strawberry$`Domain Category` <- str_trim(strawberry$`Domain Category`, side = "both")


```


```{r}
#| label: Function to remove commas and process footnotes
#| echo: false
#| warning: false
#| message: false
#| eval: true

# Function to remove commas and convert to numeric
convert_to_numeric <- function(input_vector) {
  numeric_vector <- suppressWarnings(as.numeric(str_replace_all(input_vector, ",", "")))
  return(numeric_vector)
}

# Convert 'Value' column to numeric after removing commas
strwb_census$NumericValue <- convert_to_numeric(strwb_census$Value)

# Find and process footnotes
process_footnotes <- function(value_column) {
  footnote_indices <- str_detect(value_column, "^\\([A-Z]\\)$")
  footnotes <- unique(value_column[footnote_indices])
  footnote_values <- list("Indices" = which(footnote_indices), "Footnotes" = footnotes)
  return(footnote_values)
}

footnotes_info <- process_footnotes(strwb_census$Value)

# Print first 20 values for verification
print(head(strwb_census$NumericValue, 20))
print(footnotes_info)

# Print unique values from the 'Value' column
print(unique(strwb_census$Value))

# Or, to view a random sample of 20 values from the 'Value' column
set.seed(42)  # Setting seed for reproducibility
sample_values <- sample(strwb_census$Value, 20)
print(sample_values)

```


## EDA

```{r}
library(dplyr)
library(ggplot2)

# Load the dataset
strawberry_data <- read.csv("strawberry.csv")

# Convert the 'Value' column to numeric, handling non-numeric entries gracefully
strawberry_data$Value <- as.numeric(as.character(strawberry_data$Value))

# Aggregate the strawberry production by state
strawberry_production_by_state <- strawberry_data %>%
  group_by(State) %>%
  summarise(Total_Production = sum(Value, na.rm = TRUE)) %>%
  arrange(desc(Total_Production))

# Filter out low-production states (e.g., states with Total_Production >= a threshold)
production_threshold <- 500  # Define your threshold here
strawberry_production_filtered <- strawberry_production_by_state %>%
  filter(Total_Production >= production_threshold)

# Create a bar plot to visualize the strawberry production by state with gradient fill
ggplot(strawberry_production_by_state, aes(x = Total_Production, fill = Total_Production, y = reorder(State, +Total_Production))) +
  geom_bar(stat = "identity") +
  scale_fill_gradient(low = "lightblue", high = "darkblue") +  # Gradient fill
  labs(
    title = "Total Strawberry Production by State",
    x = "State",
    y = "Total Production"
  ) +
  theme_minimal() +
  theme(axis.text.y = element_text(size = 5.75))  # Adjust the size of y-axis labels
```

California Dominates Strawberry Production: The state of California stands out as the leading producer of strawberries, with the highest total production. It is indicated by the longest horizontal bar on the chart.
While California leads in production, there is a considerable variation in production levels among other states. Some states have relatively low production, while others have moderate to high production.



## References
[NASS help](https://quickstats.nass.usda.gov/tutorials)

[Quick Stats Glossary](https://quickstats.nass.usda.gov/src/glossary.pdf)


[Quick Stats Column Definitions](https://quickstats.nass.usda.gov/param_define)



































